---
title: "PATHj computation details"
nickname: computation_details
topic: pathj
category: details
output: 
  html_document:
     includes:
         in_header: ganalytics.txt
     toc: true
     toc_float:
        collapsed: false
bibliography: 
     - bib.bib
     - bib.yaml
link-citations: true
editor_options: 
  chunk_output_type: console
---


```{r echo=FALSE,results='hide'}
source("../R/constants.R")
source("../R/functions.R")
```

`r keywords("pathj,jamovi, path analyis ,interactions,categorical variables")`

**PATHJ** module details

`r version("0.5.*")` 
`r draft`


# Introduction

`r modulename` module estimation is based on `lavaan` R package [@lavaan]. Most of `r modulename` work is to prepare data and options to be sent to lavaan, extract the output and pack it nicely for `r jamovi` output. Nonetheless, there are a few operations that should be explained to let the user know what `r modulename` is actually doing. Here you find some information.


# Categorical Variables

`r modulename` accepts any categorical variable as exogenous variable, indepenently of the number of levels in the variable. This is achieved using a very general rule of linear models: _any categorical variable can be cast in a linear model by representing it with K-1 contrast variables, where K is the number of levels_. 

Thus, when the user inserts a categorical variable in `r opt("Exogenous Factors")`, `r modulename` creates K-1 new variables in the data sets, each representing a contrast between levels. What constrast weights are used is decided by the `r opt("Factors Coding")` panel. All directives regarding the categorical variables (variances, regression coefficients, etc.) are applied to each contrast of the variable. 

Generally speaking, this method works well. Here is an example:

Consider the data `pathjexample`, present in the module. One categorical variable is called `groups_a`, featuring three groups. If one uses R `lm()` to estimate the effect of `groups_a` on `y1`, one gets the following results:

```{r}
library(pathj)
data("pathjdata")
model_formula<-y1~groups_a
pathjdata$groups_a<-factor(pathjdata$groups_a)
contrasts(pathjdata$groups_a)<-contr.treatment(3)-(1/3)  # centered constrasts with group 1 as reference group
summary(lm(model_formula,data=pathjdata))

```

If we know run the same model in `r modulename`, we get the same results in terms of coefficients and interpretation.

```{r, eval=FALSE}

mod<-pathj::pathj(formula=list(model_formula),data=pathjdata)
mod$models$r2
mod$models$coefficients

```

`r pic("pics/details/output_anova_coefficients.png")`

Standard errors are slightly different, but that is due to the fact that regression uses OLS and path analysis uses ML. We also get the same $R^2$ and significance of the $R^2$, which introduces us to the next topic.



`r issues()`

'
