---
title: "Simple models"
author: "(Marcello Gallucci)"
nickname: simplemodels
topic: pathj
category: example
output: 
  html_document:
     includes:
         in_header: ganalytics.txt
     toc: true
     toc_float:
        collapsed: false
---

```{r echo=FALSE,results='hide'}
source("../R/functions.R")
```


`r keywords("multiple regression, path analysis, lavaan, multivatiate regression")`

In this example we show how to estimate simple regression models with `r modulename`. The aim is threefold: first, it may help to start with simple model to familiarize with path analysis, and see how path analysis encompasses simple model usually estimated with the General Linear Model (but look at [GAMLj module])(https://gamlj.github.io/)). Second, we are going to estimate some models discussed in [UCLA statistical consulting webpage](https://stats.idre.ucla.edu/r/seminars/rsem/), so one can compare the results obtained in `lavaan` with the ones obtained in `r modulename`. Third, we show how to add some interesting test and twist to simple regression models.

# Research data

The data we use can be found [here at UCLA](https://stats.idre.ucla.edu/wp-content/uploads/2021/02/worland5.csv). The sample is composed by 500 students,  each with 9 observed variables: Motivation (motiv), Harmony (harm), Stability (stabi), Negative Parental Psychology (ppsych), SES, Verbal IQ (verbal), Reading (read), Arithmetic (arith) and Spelling (spell). They are all continuous variables.

# Simple regression (Model 1A)

First, we estimate a simple regression with *read* as dependent (endogenous) variable and *motiv* as independent (exogenous) variable. 

## Input

In `r modulename` we set the variables roles in the first panel as shown here:

`r pic("pics/example1/input_model1_roles.png")`

and then specify the endogeneous variable predictor in the `r opt("Endogenous Models")` panel

`r pic("pics/example1/input_model1_model.png")`

Because we want to see the path diagram of the model (although in this case is too simple to be interesting), we ask for the path diagram in the `r opt("Path Diagram")` panel.

`r pic("pics/example1/input_model1_diagram.png")`

## Output







```{r  include=FALSE}
form<-"read~motiv"
mod<-pathj::pathj(formula = form,data=data)

```

```{r  results="asis",warning=FALSE,echo=FALSE}

  cat("<pre class='jamovitable'>")
  print(modi$info)
  cat("</pre>")

```

# Related examples
`r include_examples("glm")`

`r issues()`